Modeling site-specific amino-acid preferences deepens phylogenetic estimates of the divergence of viral proteins.

### 2018-03-07 manuscript/manuscript.tex
second pass at materials and methods

We fit the GY94 model and each of the three ExpCM's to the fixed HA tree topology estimated using \texttt{RAxML}, and also tested a version of each model with $\Gamma\omega$ rate variation.
\ref{tab:empirical_data} shows that all the ExpCM's fit the actual data much better than the GY94 models.
The best fit was for the ExpCM that was informed by the average of the H1 and H3 deep mutational scans.
For all models, incorporating $\Gamma\omega$ rate variation improved the fit, although even ExpCM's without $\Gamma\omega$ greatly outperformed the GY94+$\Gamma\omega$ models.
As mentioned in the previous section, $\omega$ generally takes on values $<1$ when fit to all sites in the gene~\citep{murrell2015gene}, and our results follow this trend. 
\ref{tab:empirical_data} shows that all models that fit a single $\omega$ fit very small values, reflecting strong purifying selection against amino-acid change in HA. 
However, the ExpCM's with a single $\omega$ always fit $\omega$ values greater than the GY94 model, suggesting that the site-specific stationary state of the ExpCM's captures some of the purifying selection that the GY94 models can represent only via a small omega.
Among the models with $\Gamma\omega$, the GY94 fits all four $\omega$ categories to values $<< 1$, but the ExpCM's all fit one of the four $\omega$ categories to a value $>1$.
This increase in $\omega$ values makes sense given the different interpretation of $\omega$ in each family of models. 
The ExpCM $\omega$ is the relative rate of fixation of nonsynonymous to synonymous mutations \textit{after} accounting for the functional constraints described by the deep mutational scanning amino-acid preferences.
This more realistic null model gives ExpCM's more power to detect diversifying selection~\citep{bloom2017identification, odrigue2016detecting}, for amino-acid change, which occurs at some sites in HA due to immune selection.

@@

\begin{figure}
\centerline{\includegraphics[width=0.99\textwidth]{figures/phylobayes.pdf}}
\caption{\label{fig:phylobayes}
\textbf{Models inferred from natural sequences have similar stationary states to models defined by experimental preferences and estimate similar branch lengths.}
(A) We fit an ExpCM(H1+H3 avg)+$\Gamma\omega$ and a \modelName{MutSel} to the tree in \ref{fig:empirical_trees}. 
The \modelName{MutSel} amino-acid preferences are inferred from the natural sequences while the ExpCM amino-acid preferences are experimentally measured and then rescaled using the stringency parameter in \ref{tab:empirical_data}. 
The \modelName{MutSel} preferences are more correlated with the average of the H1 and H3 deep mutational scanning preference sets than the individual experiments are to each other (Pearson's r: 0.74 vs. 0.52). 
(B) The \modelName{MutSel} and the ExpCM(H1+H3 avg)+$\Gamma\omega$ estimated similar branch lengths. 
To aid direct comparison between the models, we normalized the branch lengths by the length between and early and late seasonal human H1 sequence. 
}
\end{figure}

The previous sections used ExpCM's, which are mutation-selection models that use site-specific amino-acid preferences that have been measured by experiments. 
However, other implementations of mutation-selection models allow their stationary states to be inferred directly from the natural sequence data. 
These models are generally implemented in a Bayesian framework, which avoids the overfitting problems associated with trying to make maximum-likelihood estimates of the thousands of parameters describing site-specific stationary states.
The model most comparable to our ExpCM's is the codon mutation-selection model implemented in \texttt{phylobayes}, which we will refer to as \modelName{MutSel}. 
The \modelName{MutSel} model is mathematically very similar to the ExpCM's except the amino-acid preferences are modeled using Dirichlet processes rather than derived from experiments. 
Like ExpCM's, \modelName{MutSel} models still assume a single set of site-specific amino-acid preferences for the entire tree.

Comparing ExpCM's and \modelName{MutSel} models can therefore help identify the ultimate limits of mutation-selection models where each site has a single time-homogeneous stationary state across the entire tree. 
If the limits of the ExpCM's reflected in Figure X are simply because the deep mutational scanning do not accurately measure the "true" amino-acid preferences of HA across the tree, then we would expect the \modelName{MutSel} models (which derive their stationary states from the entire tree) to perform better.
On the other hand, if the major limitation is simply that there is not any single set of time-homogenous amino-acid preferences that describe HA evolution over the entire tree, then we would expect \modelName{MutSel} and ExpCM models to perform similarly.

The ExpCM that estimates the longest branches is the one that uses the average of the H1 and H3 deep mutational scanning measurements with $\Gamma\omega$ rate variation~(\ref{fig:empirical_trees}).  
We therefore compared this ExpCM with a \modelName{MutSel} model given the tree in \ref{fig:empirical_trees}. 
First, the stationary state inferred by \modelName{MutSel} is similar to the average of the H1 and H3 preferences measured by deep mutational scanning. 
As \ref{fig:phylobayes}A shows, the site-specific amino-acid preferences estimated by \modelName{MutSel} are more correlated with the average amino-acid preferences from the H1 and H3 experiments than the individual deep mutational scanning experiments are to each other. 
This strong correlation between the \modelName{MutSel} and the average deep mutational scanning preference set indicates that, when averaged, the experimental preferences are able to capture the global site-specific functional constraint fairly well.

Second, the results in \ref{fig:phylobayes}B show that the ExpCM+$\Gamma\omega$(H1+H3 avg prefs) and the \modelName{MutSel} model estimated similar branch lengths on the HA tree in \ref{fig:empirical_trees}. 
While overall the two models estimated similar lengths, the tension between local and global accuracy of the amino-acid preferences is still apparent in these results. 
All the long branches from either the H1 or H3 sequences used in the experiments were estimated to be slightly longer by the ExpCM+$\Gamma\omega$, while many other branches were estimated to be slightly longer by the \modelName{MutSel} model. 
The relatively longer branches from the experimental sequences when using the ExpCM+$\Gamma\omega$ suggests that the ``global" stationary state inferred by \texttt{phylobayes} is not as accurate as the deep mutational scanning preferences for sequences close to the experimental sequences. 
However, at sequences distant from those used in the experiments, the ``global'' preferences estimated by the \modelName{MutSel} model appear to be slightly better than the average of the experimental values.
Therefore, while the site-specific \modelName{MutSel} models inferred by \texttt{phylobayes} are probably more accurate for sequences distant from the H1 and H3 in the experiments, they do not avoid the tension between the importance of a site-specific stationary state for long branch estimation and shifting preferences.

@@

\subsubsection*{Experimentally Informed Codon Model (ExpCM)}

The ExpCM is described in detail in \citet{hilton2017phydms}.

In an ExpCM, rate of substitution $P_{r,xy}$ of site $r$ from codon $x$ to $y$ is written in mutation-selection form~\citep{halpern1998evolutionary,mccandlish2014modeling,spielman2015relationship} as
\begin{equation}
P_{r,xy} = Q_{xy} \times F_{r,xy}
\end{equation}
where $Q_{xy}$ is proportional to the rate of mutation from $x$ to $y$, and $F_{r,xy}$ is proportional to the probability that this mutation fixes.
The rate of mutation $Q_{xy}$ is assumed to be uniform across sites, and takes an HKY85-like~\citep{hasegawa1985dating} form.

The deep mutational scanning data are incorporated into the ExpCM via the $F_{r,xy}$ terms.
The experiments measure the preference $\pi_{r,a}$ of every site $r$ for every amino-acid $a$.
The $F_{r,xy}$ terms are defined in terms of these experimentally measured amino-acid preferences as
\begin{equation}
\label{eq:Frxy}
F_{r,xy} = 
\begin{cases}
   1 & \mbox{if $\mathcal{A}\left(x\right) = \mathcal{A}\left(y\right)$} \\
   \omega \times \frac{\ln\left[\left(\pi_{r,\mathcal{A}\left(y\right)} / \pi_{r,\mathcal{A}\left(x\right)}\right)^{\beta}\right]}{1 - \left(\pi_{r,\mathcal{A}\left(x\right)} / \pi_{r,\mathcal{A}\left(y\right)}\right)^{\beta}} & \mbox{if $\mathcal{A}\left(x\right) \ne \mathcal{A}\left(y\right)$}
   \end{cases}
\end{equation}
where $\mathcal{A}\left(x\right)$ is the amino-acid encoded by codon $x$, $\beta$ is the stringency parameter, and $\omega$ is the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences.
The ExpCM has six free parameters (three $\phi_w$ values, $\kappa$, $\beta$, and $\omega$).
The preferences $\pi_{r,a}$ are \emph{not} free parameters since they are determined by an experiment independent of the sequence alignment being analyzed.

@@

We consider the basic Goldman-Yang style (GY94) substitution model defined as M0 in \citet{yang2000codon}. 
$P_{xy}$ is the substitution rate from codon $x$ to codon $y$ and is defined by

\begin{equation}
\label{eq:GY94}
P_xy = 
\begin{cases}
  0 & \mbox{if $x$ and $y$ differ by more than one nucleotide,}\\
  \mu \omega \Phi_{y} & \mbox{if $x$ is converted to $y$ by a single-nucleotide transversion,} \\
  \kappa \mu \omega \Phi_{y} & \mbox{if $x$ is converted to $y$ by a single-nucleotide transition,} \\
  -\sum\limits_{z \ne x} P_{xz} & \mbox{if $x = y$.}
  \end{cases}
\end{equation}
where $\kappa$ is the transition-transversion ratio, $\Phi_y$ is the equilibrium frequency of
codon $y$, $\omega$ is the gene-wide rate of non-synonymous change, and $\mu$ is the substitution rate.

\subsubsection*{GY94 with $\omega_r$}
Following~\citet{spielman2015relationship}, we inferred an $\omega$ for each site in the protein ($\omega_r$) from an ExpCM. 
\begin{equation}
\label{eq:w_r}
\omega_{r} = \frac{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times \frac{P_{r,xy}}{\omega}}}{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times Q_{xy}}}
\end{equation}
where $p_{r,x}$ is the stationary state of the ExpCM at site $r$ and codon $x$, $P_{r,xy}$ is the substitution rate from codon $x$ to codon $y$ at site $r$, $Q_{xy}$ is the mutation rate from codon $x$ to codon $y$, and $N_x$ is the set of codons that are nonsynonymous to codon $x$ and differ from codon $x$ by only one nucleotide.

We replaced the $\omega$ value in \ref{eq:GY94} with $\omega \times \omega_r$, resulting in a site-specific model with the GY94 stationary state.

\subsection*{Adding $\Gamma\omega$}

For the GY94+$\Gamma\omega$ and ExpCM+$\Gamma\omega$, $\omega$ is drawn from $K=4$ discrete categories and equal weight is given to each category. 
Set to the mean of the discretized $\Gamma$-distribution.

\subsection*{Deep Mutational Scanning Amino-Acid preferences.}
We used deep mutational scanning amino-acid preferences from an H1 HA~\citep{doud2016accurate} and an H3 HA~\citep{lee2018deep} to define the ExpCM's. 
We only considered homologous sites between the H1 and H3 protein (described in \ref{suppfile:WSN_Perth_map}). 
The resulting preference sets, including the average between the H1 and H3 preferences, are found in \ref{suppfile:H1_prefs}, \ref{suppfile:H3_prefs}, and \ref{suppfile:avg_prefs}.

@@

We downloaded all full-length, coding sequences for 14 of the 18 HA subtypes, excluding the rare subtypes 12, 15, 17, and 18. 
We filtered and aligned the sequences using \texttt{phydms\_comprehensive}~\citep{hilton2017phydms}. 
Specifically, we used \texttt{phydms\_comprehensive} default parameters except \texttt{--minidentity 0.3} and we removed sites which are not homologous between the H1 and H3 deep mutational references sequences using the sequence in \ref{suppfile:hybrid_sequence}. 
We subsampled the sequences so that there were 5 sequences per subtype with $\le 1$ sequence per year per subtype. 
%Finally, we ensured the ``landmark" sequences from the H1, H2, and H3 subtypes (\ref{suppfile:required_seqs}) were included. 
The resulting alignment contains 87 sequences and is found in \ref{suppfile:alignment_high}.

@@

We inferred the tree topology of each alignment using \texttt{RAxML}~\citep{stamatakis2006raxml} and the GTRCAT model. 
We estimated the branch lengths of given this fixed topology for the ExpCM and GY94 models using \texttt{phydms\_comprehensive}~\citep{hilton2017phydms}.

@@

We built ExpCM and GY94 models using parameters fit to the ``low divergence from H1" tree.  
Specifically, we built a GY94 model using the parameter values in \ref{suppfile:hybrid_lowH1_GY94}, a GY94+$\Gamma\omega$ model using the parameter values in \ref{suppfile:hybrid_lowH1_GY94+Gr}, an ExpCM using the parameters in \ref{suppfile:hybrid_lowH1_ExpCMdoud} and the amino-acid preferences in \ref{suppfile:H1_prefs}, and a GY94+$\omega_r$ using the GY94 parameters in \ref{suppfile:hybrid_lowH1_GY94} and the ExpCM defined by\ref{suppfile:hybrid_lowH1_ExpCMdoud} and \ref{suppfile:H1_prefs}.

For each model we calculated the expected sequence divergence given some branch length $t$ by 
\begin{equation}
\label{eq:f}
\sum_a \sum_{x \in a} p_{r,x} \sum_{y \in a} [M_{r}\left(t\right)]_{xy}
\end{equation}
where $a$ is all amino acids, $p_{r,x}$ is the stationary state of the model at site $r$ and codon $x$, and $[M_{r}\left(t\right)]_{xy}$ is the transition rate from codon $x$ to codon $y$ at site $r$ given time $t$.

@@

We simulated sequences using \texttt{pyvolve}~\citep{spielman2015pyvolve} along tree for the full alignment (\ref{fig:empirical_trees};\ref{suppfile:alignment_high}) using an ExpCM defined by the parameters in \ref{suppfile:hybrid_lowH1_ExpCMdoud} and the amino-acid preferences in \ref{suppfile:H1_prefs}. 
We performed 10 replicate simulations and estimated the branch lengths for each replicate using \texttt{phydms\_comprehensive}.

>>>>

We fit the GY94 model and each of the three ExpCM's to the fixed HA tree topology estimated using \texttt{RAxML}, and also tested a version of each model with $\Gamma\omega$ rate variation.
\ref{tab:empirical_data} shows that all the ExpCM's fit the actual data much better than the GY94 models.
The best fit was for the ExpCM that was informed by the average of the H1 and H3 deep mutational scans.
For all models, incorporating $\Gamma\omega$ rate variation improved the fit, although even ExpCM's without $\Gamma\omega$ greatly outperformed the GY94+$\Gamma\omega$ models.
As mentioned in the previous section, $\omega$ generally takes on values $<1$ when fit to all sites in the gene~\citep{murrell2015gene}, and our results follow this trend. 
\ref{tab:empirical_data} shows that all models that fit a single $\omega$ fit very small values, reflecting strong purifying selection against amino-acid change in HA. 
However, the ExpCM's with a single $\omega$ always fit $\omega$ values greater than the GY94 model, suggesting that the site-specific stationary state of the ExpCM's captures some of the purifying selection that the GY94 models can represent only via a small omega.
Among the models with $\Gamma\omega$, the GY94 fits all four $\omega$ categories to values $<< 1$, but the ExpCM's all fit one of the four $\omega$ categories to a value $>1$.
This increase in $\omega$ values makes sense given the different interpretation of $\omega$ in each family of models. 
The ExpCM $\omega$ is the relative rate of fixation of nonsynonymous to synonymous mutations \textit{after} accounting for the functional constraints described by the deep mutational scanning amino-acid preferences.
This more realistic null model gives ExpCM's more power to detect diversifying selection~\citep{bloom2017identification, rodrigue2016detecting}, for amino-acid change, which occurs at some sites in HA due to immune selection.

@@

\begin{figure}
\centerline{\includegraphics[width=0.99\textwidth]{figures/phylobayes.pdf}}
\caption{\label{fig:phylobayes}
\textbf{Models inferred from natural sequences have similar stationary states to models defined by experimental preferences and estimate similar branch lengths.}
(A) We fit an ExpCM(H1+H3 avg)+$\Gamma\omega$ and a pbMutSel to the tree in \ref{fig:empirical_trees}. 
The pbMutSel amino-acid preferences are inferred from the natural sequences while the ExpCM amino-acid preferences are experimentally measured and then rescaled using the stringency parameter in \ref{tab:empirical_data}. 
The pbMutSel preferences are more correlated with the average of the H1 and H3 deep mutational scanning preference sets than the individual experiments are to each other (Pearson's r: 0.74 vs. 0.52). 
(B) The pbMutSel and the ExpCM(H1+H3 avg)+$\Gamma\omega$ estimated similar branch lengths. 
To aid direct comparison between the models, we normalized the branch lengths by the length between and early and late seasonal human H1 sequence. 
}
\end{figure}

The previous sections used ExpCM's, which are mutation-selection models that use site-specific amino-acid preferences that have been measured by experiments. 
However, other implementations of mutation-selection models allow their stationary states to be inferred directly from the natural sequence data. 
These models are generally implemented in a Bayesian framework, which avoids the overfitting problems associated with trying to make maximum-likelihood estimates of the thousands of parameters describing site-specific stationary states.
The model most comparable to our ExpCM's is the codon mutation-selection model implemented in \texttt{phylobayes}, which we will refer to as pbMutSel. 
The pbMutSel model is mathematically very similar to the ExpCM's except the amino-acid preferences are modeled using Dirichlet processes rather than derived from experiments. 
Like ExpCM's, pbMutSel models still assume a single set of site-specific amino-acid preferences for the entire tree.

Comparing ExpCM's and pbMutSel models can therefore help identify the ultimate limits of mutation-selection models where each site has a single time-homogeneous stationary state across the entire tree. 
If the limits of the ExpCM's reflected in Figure X are simply because the deep mutational scanning do not accurately measure the "true" amino-acid preferences of HA across the tree, then we would expect the pbMutSel models (which derive their stationary states from the entire tree) to perform better.
On the other hand, if the major limitation is simply that there is not any single set of time-homogenous amino-acid preferences that describe HA evolution over the entire tree, then we would expect pbMutSel and ExpCM models to perform similarly.

The ExpCM that estimates the longest branches is the one that uses the average of the H1 and H3 deep mutational scanning measurements with $\Gamma\omega$ rate variation~(\ref{fig:empirical_trees}).  
We therefore compared this ExpCM with a pbMutSel model given the tree in \ref{fig:empirical_trees}. 
First, the stationary state inferred by pbMutSel is similar to the average of the H1 and H3 preferences measured by deep mutational scanning. 
As \ref{fig:phylobayes}A shows, the site-specific amino-acid preferences estimated by pbMutSel are more correlated with the average amino-acid preferences from the H1 and H3 experiments than the individual deep mutational scanning experiments are to each other. 
This strong correlation between the pbMutSel and the average deep mutational scanning preference set indicates that, when averaged, the experimental preferences are able to capture the global site-specific functional constraint fairly well.

Second, the results in \ref{fig:phylobayes}B show that the ExpCM+$\Gamma\omega$(H1+H3 avg prefs) and the pbMutSel model estimated similar branch lengths on the HA tree in \ref{fig:empirical_trees}. 
While overall the two models estimated similar lengths, the tension between local and global accuracy of the amino-acid preferences is still apparent in these results. 
All the long branches from either the H1 or H3 sequences used in the experiments were estimated to be slightly longer by the ExpCM+$\Gamma\omega$, while many other branches were estimated to be slightly longer by the pbMutSel model. 
The relatively longer branches from the experimental sequences when using the ExpCM+$\Gamma\omega$ suggests that the ``global" stationary state inferred by \texttt{phylobayes} is not as accurate as the deep mutational scanning preferences for sequences close to the experimental sequences. 
However, at sequences distant from those used in the experiments, the ``global'' preferences estimated by the pbMutSel model appear to be slightly better than the average of the experimental values.
Therefore, while the site-specific pbMutSel models inferred by \texttt{phylobayes} are probably more accurate for sequences distant from the H1 and H3 in the experiments, they do not avoid the tension between the importance of a site-specific stationary state for long branch estimation and shifting preferences.

@@

\subsubsection*{Experimentally Informed Codon Model (ExpCM).}

In an ExpCM, the rate of substitution $P_{r,xy}$ of site $r$ from codon $x$ to $y$ is written in mutation-selection form~\citep{halpern1998evolutionary,mccandlish2014modeling,spielman2015relationship} as
\begin{equation}
\label{eq:ExpCM}
P_{r,xy} = Q_{xy} \times F_{r,xy}
\end{equation}
where $Q_{xy}$ is proportional to the rate of mutation from $x$ to $y$, and $F_{r,xy}$ is proportional to the probability that this mutation fixes.
The rate of mutation $Q_{xy}$ is assumed to be uniform across sites, and takes an HKY85-like~\citep{hasegawa1985dating} form.

The deep mutational scanning amino-acid preferences are incorporated into the ExpCM via the $F_{r,xy}$ terms.
The experiments measure the preference $\pi_{r,a}$ of every site $r$ for every amino-acid $a$.
The $F_{r,xy}$ terms are defined in terms of these experimentally measured amino-acid preferences as
\begin{equation}
\label{eq:Frxy}
F_{r,xy} = 
\begin{cases}
   1 & \mbox{if $\mathcal{A}\left(x\right) = \mathcal{A}\left(y\right)$} \\
   \omega \times \frac{\ln\left[\left(\pi_{r,\mathcal{A}\left(y\right)} / \pi_{r,\mathcal{A}\left(x\right)}\right)^{\beta}\right]}{1 - \left(\pi_{r,\mathcal{A}\left(x\right)} / \pi_{r,\mathcal{A}\left(y\right)}\right)^{\beta}} & \mbox{if $\mathcal{A}\left(x\right) \ne \mathcal{A}\left(y\right)$}
   \end{cases}
\end{equation}
where $\mathcal{A}\left(x\right)$ is the amino-acid encoded by codon $x$, $\beta$ is the stringency parameter, and $\omega$ is the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences.
The preferences $\pi_{r,a}$ are \emph{not} free parameters since they are determined by an experiment independent of the sequence alignment being analyzed.
The ExpCM is described in detail in~\citet{hilton2017phydms}.

@@

The GY94 model is defined as the M0 in \cite{yang2000codon}. 
$P_{xy}$ is the substitution rate from codon $x$ to codon $y$ and is defined by

\begin{equation}
\label{eq:GY94}
P_{xy} = 
\begin{cases}
  0 & \mbox{if $x$ and $y$ differ by more than one nucleotide,}\\
  \mu \omega \Phi_{y} & \mbox{if $x$ is converted to $y$ by a single-nucleotide transversion,} \\
  \kappa \mu \omega \Phi_{y} & \mbox{if $x$ is converted to $y$ by a single-nucleotide transition,} \\
  -\sum\limits_{z \ne x} P_{xz} & \mbox{if $x = y$.}
  \end{cases}
\end{equation}
where $\kappa$ is the transition-transversion ratio, $\Phi_y$ is the equilibrium frequency of
codon $y$, $\omega$ is the gene-wide rate of non-synonymous change, and $\mu$ is the substitution rate.

\subsubsection*{GY94 with $\omega_r$.}

We defined a site-specific model with the GY94 stationary state by multiplying the $\omega$ in \ref{eq:GY94} with a site-specific $\omega_r$ value calculated from an ExpCM (\ref{eq:ExpCM}). 
Specifically, following \citet{spielman2015relationship}, 
\begin{equation}
\label{eq:w_r}
\omega_{r} = \frac{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times \frac{P_{r,xy}}{\omega}}}{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times Q_{xy}}}
\end{equation}
where $p_{r,x}$ is the stationary state of the ExpCM at site $r$ and codon $x$, $P_{r,xy}$ is the substitution rate from codon $x$ to codon $y$ at site $r$, $Q_{xy}$ is the mutation rate from codon $x$ to codon $y$, $\omega$ is the relative nonsynonymous substitution rate from the ExpCM and $N_x$ is the set of codons that are nonsynonymous to codon $x$ and differ from codon $x$ by only one nucleotide.

\subsection*{$\Gamma\omega$ rate variation.}

The GY94+$\Gamma\omega$ is defined as M5 in \citet{yang2000codon} and the ExpCM+$\Gamma\omega$ follows the same logic. 
Briefly, $\omega$ is drawn from a $\Gamma$ distribution discretized into $K=4$ bins. 
Each bin is equally weighted and $\omega$ takes on the mean value of the bin.

\subsection*{Deep mutational scanning amino-acid preferences.}
We used amino-acid preferences measured in deep mutational scanning of an H1 HA~\citep{doud2016accurate} and an H3 HA~\citep{lee2018deep} to define the $\pi_{r, A\left(x\right)}$ parameters~(\ref{eq:Frxy}) of the ExpCM's. 
We only considered homologous sites between the H1 and H3 protein, described in \ref{suppfile:WSN_Perth_map}. 
The resulting preference sets, including the average between the H1 and H3 preferences, are found in \ref{suppfile:H1_prefs}, \ref{suppfile:H3_prefs}, and \ref{suppfile:avg_prefs}.

@@

We downloaded all full-length, coding sequences for 14 of the 18 HA subtypes, excluding the rare subtypes 12, 15, 17, and 18. 
We filtered and aligned the sequences using \texttt{phydms\_prepalignment}~\citep{hilton2017phydms}. 
Specifically, we used \texttt{phydms\_prepalignment} with the flag \texttt{--minidentity 0.3} to sequences with ambiguous nucleotides, premature stops, or frameshift mutations as well as redundant sequences.  
We also removed any codon sites which are not shared between the H1 and H3 deep mutational scanning reference sequences by aligning to sequence in \ref{suppfile:hybrid_sequence}. 
We subsampled the remaining sequences so there are $\ge 5$ sequences per subtype with $\le 1$ sequence per year per subtype. 
The resulting alignment contains 87 sequences and is found in \ref{suppfile:alignment_high}.

@@

We inferred the tree topology of each alignment using \texttt{RAxML}~\citep{stamatakis2006raxml} and the GTRCAT model. 
We estimated the branch lengths of this fixed topology using the ExpCM and GY94 models with \texttt{phydms\_comprehensive}~\citep{hilton2017phydms}.

@@

We defined ExpCM and GY94 models using parameter values fit to the ``low divergence from H1" tree~(\ref{fig:empirical_trees}, \ref{suppfile:alignment_lowH1}). 
Specifically, we defined the GY94 model using the parameter values in \ref{suppfile:hybrid_lowH1_GY94}, the GY94+$\Gamma\omega$ using the parameter values in\ref{suppfile:hybrid_lowH1_GY94+Gr} and the ExpCM using the parameter values in \ref{suppfile:hybrid_lowH1_ExpCMdoud} and the H1 amino-acid preferences from~\citet{doud2016accurate} in \ref{suppfile:H1_prefs}.

For each model, we calculated the expected sequence divergence given branch length $t$ by 
\begin{equation}
\label{eq:f}
\sum_a \sum_{x \in a} p_{r,x} \sum_{y \in a} [M_{r}\left(t\right)]_{xy}
\end{equation}
where $a$ is all amino acids, $p_{r,x}$ is the stationary state of the model at site $r$ and codon $x$, and $[M_{r}\left(t\right)]_{xy}$ is the transition rate from codon $x$ to codon $y$ at site $r$ given time $t$.

@@

We simulated sequences using \texttt{pyvolve}~\citep{spielman2015pyvolve} along the tree for the full alignment (\ref{fig:empirical_trees}, \ref{suppfile:alignment_high}) using an ExpCM defined by parameters fit to the ``low divergence from H1 tree" (\ref{suppfile:alignment_lowH1}). 
Specifically, we defined the ExpCM using the parameter values in \ref{suppfile:hybrid_lowH1_ExpCMdoud} and the H1 amino-acid preferences from \citet{doud2016accurate} in \ref{suppfile:H1_prefs}. 
We performed 10 replicate simulations and estimated the branch lengths for each replicate using \texttt{phydms\_comprehensive}~\citep{hilton2017phydms}.

\subsection*{pbMutSel inference with \texttt{phylobayes}.}

We inferred a pbMutSel model for the full tree (\ref{fig:empirical_trees}, \ref{suppfile:alignment_high}). 
We ran one chain for 5000 steps, saved every sample, and discarded the first 500 samples as a burnin. 
We normalized the branch lengths on the pbMutSel consensus tree and the ExpCM(H1+H3 avg)+$\Gamma\omega$ by dividing each branch by the length from A/South Carolina/1/1918 and A/Solomon Islands/3/2006. 
These two H1 sequences are early and late representatives of the longest continuous sampled \skhcomment{in evolution but not in time...} lineage of human influenza.

\subsection*{Code and analysis.}
All code used is available at \url{https://github.com/jbloomlab/divergence_timing_manuscript}. 
For the main analysis, we used \texttt{phydms}~\citep{hilton2017phydms} version 2.2.dev2 (available at \url{github.com/jbloomlab/phydms}) to filter sequences and estimate branch lengths, \texttt{pyvolve}~\citep{spielman2015pyvolve} version 0.8.7 (available at \url{https://github.com/sjspielman/pyvolve}) to simulate the sequences and \texttt{phylobayes}~\citep{rodrigue2014site} version 1.8 (available at \url{https://github.com/bayesiancook/pbmpi}) to infer the pbMutSel model.

We also used \texttt{ggplot2}~\citep{wickham2016ggplot2}, \texttt{ggtree}~\citep{yu2017ggtree}, and \texttt{ggseqlogo}~\citep{wagih2017ggseqlogo} for visualization.
